% !TEX encoding = UTF-8 Unicode
% !TEX program = pdflatex
% !TEX spellcheck = en_US


% In order to correctly compile this document,
% execute the following commands:
% 1. pdflatex
% 2. pdflatex
% 3. pdflatex



\documentclass[amsthm,ebook]{saparticle}

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

\usepackage{teubner}
% IF YOU USE XELATEX
%\usepackage{polyglossia}
% if you write in italian
%\setmainlanguage{italian}
% If you want put some ancient greek:
%\setotherlanguage[variant=polytonic]{greek}
%\newfontfamily{\greekfont}[Ligatures=TeX]{Palatino Linotype}

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Where Eagles Dare}
%\author[]{Werner Eck\corref{first}}
\author[kcl]{Charlotte Roueché\corref{first}}
\address[kcl]{King's College London} 
\cortext[first]{Corresponding author. Email: charlotte.roueche@kcl.ac.uk}
\begin{document}
\maketitle
%\begin{abstract}
%\end{abstract}

I couldn't resist this title, taken from Shakespeare's \textit{Richard III} to 
use for a Hollywood film, full of daring wartime adventures. As academics we may 
underestimate the importance of courage in our undertakings. But the early epigraphers 
required a good deal of \textbf{daring}; while recording inscriptions was straightforward 
enough in Italy or north-western Europe, the eastern and southern regions of the 
Greco-Roman world remained difficult and dangerous to visit until the late nineteenth 
century or even later – as is again becoming true. The early travellers also 
suffered from practical constraints; in the 18\textsuperscript{th} and 19\textsuperscript{th} centuries 
they were limited by the quantity of paper that they had brought with them just 
as in the 20\textsuperscript{th} century; Robert `Palmyra' Wood recorded inscriptions 
on pages of his copy of Homer. He was travelling for several months, and was saving 
his paper for sketches, plans and drawings of buildings\footnote{Robert Wood (1716/17–1771): doi:10.1093/ref:odnb/29891}. Similarly John Deering 
recorded texts in a notebook with very small pages\footnote{John Peter Deering (1787–1850): doi:10.1093/ref:odnb/7420}. Twentieth century travellers 
were to experience similar constraints on the amount of photographic film available 
to them.

At the same time, epigraphers have always been ingenious in their use of technical 
solutions. The most dramatic of these is perhaps the development of \textbf{squeezes}, 
which have turned out to provide records of enduring and continuing value. The 
driving force here was the need to record inscriptions in languages – hieroglyphics, 
cuneiform – which could not be interpreted, and even incised designs; copyists 
who know Greek or Latin could record texts in those languages with relative ease, 
but they too came to find squeezes useful in representing what they saw. Over time 
the technology was improved; while the plaster casts of Egyptian paintings could 
seriously damage the monument, paper squeezes improved in quality\footnote{\url{http://www.asia.si.edu/research/squeezeproject/sq_making.asp}}. 

During the 19th century, \textbf{travel} became steadily safer, and also easier 
- most dramatically as railways began to open up new regions. Over the same period, 
the creation of the first great corpora of inscriptions encouraged an increasing 
standardisation of records. Early travellers recorded as much as they could, often 
in haste: this normally took the form of a simple transcription, or drawing, of 
the text, with sometimes a brief mention of its support.  Gradually, measurements 
start to appear – often only of letters, then of the monument or fragment itself. 
In 1890 the Austrian government set up the Kleinasiatische Kommission\footnote{\url{http://www.oeaw.ac.at/en/science-and-society/commissions/kleinasiatische-kommission/geschichte/}}, which provided 
travellers with special notebooks: these were pre-printed with headings for Location 
and Position, Material, Height, Width, Depth, Letter heights, Shape and Condition, 
Number and location of squeeze, When copied and by whom\footnote{Fund- und Standort, Material, Höhe, Breite, Dicke, Buchstabenhöhe, Form und Erhaltung, Nummer und Ortsangabe des Abklatsches, Wann und von wem copirt}. On return to Vienna, the 
notebooks could be filed, and the squeezes stored, in bookshelves and drawers specially 
designed for the purpose.

At the same time, the publication of the corpora, and the great projects from Boeckh's 
\textit{CIG} to  Mommsen's \textit{CIL} was revealing the \textbf{volume} of material. 
 The Ottoman world was becoming increasingly accessible; in the western part of 
the Roman Empire development and industrialisation – particularly the redevelopment 
of Rome as a capital city -- increased the torrent of material. Publishing an abundance 
of texts, accompanied by increasingly detailed information, required a systematised 
response. Organisation could be thematic: Christian inscriptions, for example, 
were identified as a separate category, requiring different expertise, although 
this division has remained problematic. It could be geographic: the Berlin Academy 
took responsibility for publishing the material from Italy and the west, while 
the Austrian Academy was to deal with Asia Minor and the East; but national interests 
also played a part, with Italian and French scholars publishing materials from 
the epigraphically prolific north African regions which their governments controlled. 
The situation also demanded finding aids – the \textit{PIR} can be seen as a 
tool for accessing the material in \textit{CIL}. And the pressure for standardisation 
continued, although it was not until the 1930s that use of the Leiden conventions 
for publishing inscribed texts was agreed (and modifications continued)\footnote{B.A. van Groningen, ‘Projet d’unification des systèmes de signes critiques’, \emph{Chronique d’Égypte}, 7 (1932): 262–9.7 ; for further modifications, see S.Dow, \emph{Conventions in Editing: A Suggested Reformulation of the Leiden System} (Durham, NC: Duke University, 1969); H. Krummrey and S. Panciera, ‘Criteri di edizione e segni diacritici’, Tituli, 2 (1980. 205-15; S. Panciera, ‘Struttura dei supplementi e segni diacritici dieci anni dopo’, \emph{Supplementa Italica} 8 (Rome: Quasar, 1991), pp. 9–21.}. 

Much of this reflected a response to the increased volume of material becoming 
available as the world changed. But the end of the nineteenth century saw the beginning 
of a further technological revolution with the arrival of \textbf{photography}. 
The use of cameras for archaeological records was at first limited to established 
excavation sites, or cities such as Athens, with a reasonable amount of infrastructure 
and protection for cumbersome equipment; but by the 1920s cameras were sufficiently 
portable to be taken out into the countryside. More and more inscriptions were 
photographed; but the traditions which had already developed meant that they were 
not immediately seen as essential elements in publication. The \textit{Monumenta 
Asiae Minoris Antiqua} represent an honourable exception, established with the 
specific aim of taking and publishing photographs; but they had the support of 
American funding\footnote{C. Roueche, `The history of an idea : tracing the origins of the MAMA project.' \emph{Roman Phrygia : Culture and Society}. ed. P. Thonemann (Cambridge, 2013),  249-264.}.  

Gradually, however, photographs began to effect transformations in scholarly practice. 
They permitted a far better understanding of regional, cultural and chronological 
distinctions, which could be communicated to readers who would have no opportunity 
to see the stones themselves. And they provided an increasing understanding of 
\textbf{context} – whose importance was emphasised by both Louis Robert and, 
more recently, Werner Eck. The photograph can present the support, and, when applicable, 
the setting of that support; Robert emphasised the importance of visualising the 
landscape surrounding a particular community.

Robert also demanded ever higher \textbf{standards} in the accompanying commentaries 
on inscriptions. But all of this raised a huge logistical problem: a text accompanied 
by a detailed description, a detailed commentary and one or more photographs requires 
a good deal of space – and more and more texts were appearing. From the 1980s 
onwards it was also becoming standard to provide a translation into a modern language. 
Publication in book form was becoming increasingly expensive and burdensome.

It was therefore changes which had been brought about by a series of technological 
developments which led far-sighted scholars – in particular Silvio Panciera  
and Geza Alföldy – to look to yet another technology. As early as the 1980s 
they both saw the value of \textbf{computers} as tools for holding, organising 
and searching large volumes of text; others quickly followed. Panciera also understood 
early on that working in this medium required collaboration, and the use of agreed 
standards, convening meetings to discuss such matters from 1989 onwards.  With 
the arrival of the web, and the resultant possibilities for communication, these 
requirements became ever more important; in the early 2000s the agreed conventions 
of epigraphy were translated into a set of machine readable instructions by Tom 
Elliott, when he developed EpiDoc\footnote{\url{http://epidoc.sourceforge.net/}}. 

At the same time, the steady improvements in technology were making  it possible, 
by the 2000s, to exchange images as well as texts; and the arrival of the digital 
camera was transforming the possibilities for photographing texts. The epigrapher 
in the field no longer depends on a finite supply of film: the traditional shot 
of several fragments photographed together for reasons of economy is disappearing. 
Instead, the epigrapher should be expected to present images of every side of a 
monument and its setting. All these developments both enabled and necessitated 
the first large scale publications of inscriptions on line: \textit{Vindolanda 
Tablets Online}\footnote{\url{http://vindolanda.csad.ox.ac.uk/}}, (2003), the \textit{U.S. Epigraphy project}\footnote{\url{http://usepigraphy.brown.edu}}, (2003–), \textit{Aphrodisias 
in Late Antiquity} (2004)\footnote{\url{http://insaph.kcl.ac.uk/ala2004/}},\textsuperscript{ }  the \textit{Inscriptions of Aphrodisias} 
(2007)\footnote{\url{http://insaph.kcl.ac.uk/iaph2007/}}, and others in preparation.

While we were working on the materials from Aphrodisias, new possibilities were 
opening up: connections were becoming faster and more ubiquitous. More and more 
relevant material was being published online: what Tim Berners-Lee calls t\label{HGoBack}\textit{he 
next Web of open}\textit{\textbf{, linked data. }}In 2008 we received a grant to 
start exploring the use of geodata with inscriptions, in the \textit{Inscriptions 
of Roman Tripolitania} (2009)\footnote{\url{http://inslib.kcl.ac.uk/irt2009/}}; this approach has been further developed in \textit{Monumenta 
Asiae Minoris Antiqua} XI\footnote{\url{http://mama.csad.ox.ac.uk/}},  and other kinds of interconnection are being actively 
explored. Linking is closely connected to \textbf{sharing}: it is becoming increasingly 
clear that one way of ensuring the survival of our materials is by making them 
openly available for others to reuse and share as widely as possible.

One of the aims of both Alföldy and Panciera had been to develop collaborative 
corpora, places where large quantities of texts could be shared to enable extensive 
searching. After their projects moved onto the web it also became possible to include 
more and more material – photographs and other images, and geodata. It was Silvia 
Orlandi who realised that the next step was a portal, to offer access across these 
and many other online epigraphic collections. Work over several years, by many 
different scholars, had established EpiDoc as a robust system, particularly for 
the exchange of information, and it was therefore available to build \textbf{EAGLE}. 
The spirit of this enterprise was exactly the spirit behind Europeana –  a project 
to present high quality records of heritage materials to a worldwide audience. 

This is an account, therefore, of an academic discipline which has evolved by engagement 
with a series of technological developments over two centuries, and is continuing 
to do so; it is also a story of developing steadily higher standards for the publication 
of heritage materials. The current challenge is to confront the fact that such 
materials will now be universally available, and must therefore be presented in 
a way that helps and supports \textbf{new audiences}. For this the EAGLE project 
has been developing valuable new resources – such as the mobile app –and, very 
importantly, encouraging translations. The crucial thing to realise is that it 
will not be possible to revert to earlier models: this project sets new, higher 
standards for epigraphic publication. This is a project which will take the subject 
into the future `on eagles' wings', as the Bible puts it – in another phrase 
used by Hollywood\footnote{Exodus 19.4.}.

They should be taking us all with them.

\end{document}


